The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-02-24 src/ms.tex
spelling and fixes
Signed-off-by: nstarman <nstarkman@protonmail.com>

The relationship between \astropy and community is reciprocal, with the
growing and evolving needs of the community and the maturation of the broader
scientific \python ecosystem now driving the development of the \astropypkg
core package.

One manifestation of this was the realization that some features do not actually
belong in the core package, either because they are of more general use
outside of \astropypkg, or because they are too specialized. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
\astropy carried, as well as the \python wrappers for it.  Since this is more
basic infrastructure, with a release cycle set by SOFA, it made more sense to
create separate packages that \astropypkg can depend on: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
was too specialized was a Virtual Observatory sub-package (now removed).  Of
its content, the Simple Application Message Protocol made sense to keep
(\astropysubpkg{samp}), but other parts belonged more properly in {\tt
Astroquery} (see Sect.~\ref{sec:astroquery}).

@@

% \begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
%   \item ERFA module is now it’s own thing; Obliquely tie in that units
%    remains in astropy -> above
%  \item ? Pytest plugin ecosystem -> done elsewhere
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
%  \item WCS API : A shared Python interface for World Coordinate Systems
%        Include APE 14 reference
% \end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the five main new features that entered the astropy core since the
previous summary \citepalias{astropy:2018}.

@@

\paragraph{Uncertainties and Distributions.} Early in our carreers, we
were likely all taught that a measurement without a unit or an
uncertainty is essentially meaningless.  \astropyunits already
provides the ability to associate numbers with units, and propagate
these correctly, with its \astropyQuantity\ class, but so far the
ability to associate and propagate uncertainties is relatively
limited: in \astropysubpkg{nddata}, there are options to associate
errors with data arrays, but these are treated as independent, i.e.,
covariances are not tracked.

In general, proper error propagation is difficult, and often it is
best to use Monte Carle methods.  With the new
\astropysubpkg{uncertainties} sub-package, a first step is made to
make this possible.  It allows one to generate, for each variable,
randomly drawn samples in a \astropyDistribution, and then propagate
these by passing them through the normal analysis, producing a
\astropyDistribution\ of final results that can be inspected.

This \astropyDistribution\ can be any type of array, including of
course a \astropyQuantity, and the goal is to ensure it can be
seamlessly used to instantiate also other astropy classes, such as
\astropySkyCoord\ and \astropyTime.

@@

    Numerous \python libraries have custom table implementations
    \begin{itemize}
      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
      \item Converting SkyCoord to QTable (1 sentence addition to above)
      \item Added support for dask arrays in tables
      \item Added support for registering array-like objects as mixin columns
    \end{itemize}

@@

  % MHvK: maybe do this before the structure? Change in structure
  % partially driven by this?
    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

@@

    For reproducibility, \astropypkg allows the constants (and therefore units)
    definitions to be rolled back to prior values, for instance the
    \texttt{SI/CODATA 2014} definitions.
    % CITATION
    The constants and units definitions are set by configuration file as well as
    \astropyScienceState (an runtime configuration controller). For work where
    updates to the values of the fundamental constants significantly alters
    results we recommend including with the work an astropy configuration file,
    which specifies the set of constants used.

@@

    \astropycosmology contains classes for representing cosmological models.
    % currently all homogeneous and isotropic variations of the
    % Friedmann-Lemaitre-Robertson-Walker metric.
    Bundled with the classes are commonly used \astropyCosmology realizations, for
    example best-fit measurements from
    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
    \citep{PlanckMission:2006}.

@@

    Being able to describe the position of objects, i.e. their coordinates, is
    fundamental across many knowledge domains. \astropypkg provides
    \astropycoordinates for working with low-level positional and velocity data
    all the way to high-level coordinate objects with reference frames,
    atmospheric information, and more. A central feature of \astropycoordinates is
    the ability to transform data between reference frames, e.g. \astropyICRS
    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
    there have been some historical limitations for spatially proximate
    transformations, impacting the usefulness of \astropycoordinates for
    ground-based telescopes and intra-solar astronomy. Transformations in the
    \astropyAltAz frame were reasonably precise for very distant objects, but
    wrong by up around several arcseconds for e.g. the location of the moon. Now
    these transformations are much more precise, down to the milli-arcsecond
    level. Similar precision improvements were made to the
    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
    transformations. Additionally,
    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
    and associated transformations have been updated to correctly reflect the
    “true” (includes precession but not nutation) and “mean” (precession and
    nutation) terminology.

@@

% \subsection*{Future} \label{sec:core-features-future}
% MHvK: Merged with the new feature section

>>>>

  The relationship between \astropy and community is reciprocal, with the
  growing and evolving needs of the community and the maturation of the broader
  scientific \python ecosystem driving much of the development of the
  \astropypkg core package.

  One manifestation of this was the realization that some features do not
  actually belong in the core package, either because they are of more general
  use outside of \astropypkg, or because they are too specialized. An example of
  the former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
  \astropy carried, as well as the \python wrappers for it.  Since this is more
  basic infrastructure, with a release cycle set by SOFA, it made more sense to
  create separate packages that \astropypkg can depend on: \package{ERFA}
  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
  was too specialized was a Virtual Observatory sub-package (now removed).  Of
  its content, the Simple Application Message Protocol made sense to keep
  (\astropysubpkg{samp}), but other parts belonged more properly in \package{
  astroquery} (see Sect.~\ref{sec:astroquery}).

@@

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the five main new features that entered the \astropypkg core since the
previous summary \citepalias{astropy:2018}.

@@

\paragraph{Uncertainties and Distributions.} Early in our careers, we
were likely all taught that a measurement without a unit or an
uncertainty is essentially meaningless.  \astropyunits already
provides the ability to associate numbers with units, and propagate
these correctly, with its \astropyQuantity\ class, but so far the
ability to associate and propagate uncertainties is relatively
limited: in \astropysubpkg{nddata}, there are options to associate
errors with data arrays, but these are treated as independent, i.e.,
covariances are not tracked.

In general, proper error propagation is difficult, and often it is
best to use Monte Carlo methods.  With the new
\astropysubpkg{uncertainties} sub-package, a first step is made to
make this possible.  It allows one to generate, for each variable,
randomly drawn samples in a \astropyDistribution, and then propagate
these by passing them through the normal analysis, producing a
\astropyDistribution\ of final results that can be inspected.

This \astropyDistribution\ can be any type of array, including of
course a \astropyQuantity, and the goal is to ensure it can be
seamlessly used to instantiate also other \astropypkg classes, such as
\astropySkyCoord\ and \astropyTime.

@@

    Numerous \python libraries have custom table implementations
    \begin{itemize}
      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
      \item Converting SkyCoord to QTable (1 sentence addition to above)
      \item Added support for \package{dask} arrays in tables
      \item Added support for registering array-like objects as mixin columns
    \end{itemize}

@@

  % MHvK: maybe do this before the structure? Change in structure
  % partially driven by this?
    The improvements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

@@

    For reproducibility, \astropypkg allows the constants (and therefore units)
    definitions to be rolled back to prior values, for instance the
    \texttt{SI/CODATA 2014} definitions.
    % CITATION
    The constants and units definitions are set by configuration file as well as
    \astropyScienceState (an runtime configuration controller). For work where
    updates to the values of the fundamental constants significantly alters
    results we recommend including with the work an \astropy configuration file,
    which specifies the set of constants used.

@@

    \astropycosmology contains classes for representing cosmological models.
    Bundled with the classes are commonly used \astropyCosmology realizations, for
    example best-fit measurements from
    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
    \citep{PlanckMission:2006}.

@@

    Being able to describe the position of objects, i.e. their coordinates, is
    fundamental across many knowledge domains. \astropypkg provides
    \astropycoordinates for working with low-level positional and velocity data
    all the way to high-level coordinate objects with reference frames,
    atmospheric information, and more. A central feature of \astropycoordinates is
    the ability to transform data between reference frames, e.g. \astropyICRS
    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
    there have been some historical limitations for spatially proximate
    transformations, impacting the usefulness of \astropycoordinates for
    ground-based telescopes and intra-solar astronomy. Transformations in the
    \astropyAltAz frame were reasonably precise for very distant objects, but
    wrong by up around several arcseconds for e.g. the location of the moon. Now
    these transformations are much more precise, down to the milliarcsecond
    level. Similar precision improvements were made to the
    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
    transformations. Additionally,
    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
    and associated transformations have been updated to correctly reflect the
    “true” (includes precession but not nutation) and “mean” (precession and
    nutation) terminology.

@@
